A quantum parallel Markov chain Monte Carlo

### 2023-01-25 manuscript/main.tex
just submitted revision

 \begin{figure}[!t]
 	\vspace{3em}
	\centering
	\includegraphics[width=0.9\linewidth]{mcmcIterations.pdf}
	\caption{Total number of oracle evaluations required for each of 2,000 parallel MCMC iterations. In total, the combined MCMC runs required roughly $7$\% of the usual 20 million target evaluations. Over  99.4\% of the 10,000 MCMC iterations successfully sampled from the discrete distribution with probabilities of \eqref{eq:probs}. Burn-in iterations require moderately more evaluations because the current state occupies a lower density region and represents a `less good' warm-start.}\label{fig:mcmcIterations}
\end{figure}

@@

In general MCMC, one often calibrates the scaling of a proposal kernel to balance between exploring the target's domain and remaining within high-density regions.  Optimal scaling strategies may lead to a large number of rejected proposals \citep{roberts2001optimal}.  Indeed, \citet{holbrook2021generating} shows that parallel MCMC algorithms are no exception. When using the Gumbel-max trick to sample from proposals, this means that the current state is often quite near optimality, representing a warm-start.  Figure \ref{fig:mcmcIterations} shows how this warm-starting effects the number of oracle calls (and hence target evaluations) within the quantum minimization sub-routine over the course of an MCMC trajectory.  We target multi-dimensional standard normal distributions of differing dimensions using the vector $(100,\dots,100)$ as starting position. We fix the number of Gaussian proposals to be 2,000 and use standard adaptation procedures to target a 50\% acceptance rate while guaranteeing convergence \citep{rosenthal2011optimal}. Across iterations, QPMCMC requires relatively few oracle calls compared to the 2,000 target evaluations of parallel MCMC. We also witness the expected drop in the number of oracle evaluations as the chain approaches the target's high-density region. Remarkably, the algorithm succeeds in sampling from the true discrete distribution in 99.5\% of the MCMC iterations. An independent simulation obtains similar results, requiring roughly 7\% of the usual number of target evaluations. Figure \ref{fig:qq} shows a quantile-quantile plot for all 100 dimensions of a multi-dimensional standard normal distribution. We see no appreciable impact from the rare failure of the quantum minimization sub-routine.

@@

\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}
\ra{1.2}
\begin{table}[t!]
	\centering
	\resizebox{\textwidth}{!}{
		\begin{tabular}{@{}lllll@{}}\toprule
			Proposals & MCMC iterations  & Target evaluations & Speedup  & Efficiency gain  \\   \midrule
			1,000 & 249,398 (200,998,  311,998) & 24,988,963 (20,149,132, 31,265,011) & 9.98  (9.98, 9.98) & 1  \\   
			5,000 & 14,398 (12,998, 16,998) & 3,314,560 (2,989,418,  3,916,281) & 21.72  (21.70, 21.74) & 7.58  (6.25, 9.71) \\   
			10,000 & 5,998  (4,998,  6,998) & 1,993,484  (1,662,592,  2,330,842) & 30  (29.96, 30.26) & 12.87  (8.64, 18.80) \\    \bottomrule
		\end{tabular}
	}
	\caption{Racing to 100 effective samples for a target with 1,000 disjoint modes. `Speedup' is ratio between target evaluations required for sequential and quantum implementations. `Efficiency gain' is ratio between target evaluations required for 1,000 proposal and 5,000/10,000 proposal implementations. We report means (minima, maxima) across 5 independent runs.}\label{tab:res}
\end{table}

@@

\section{Note on simulations}

We use \textsc{R} \citep{rlang}, \textsc{Python} \citep{vanrossum1995python}, \textsc{TensorFlow} \citep{abadi2016tensorflow} and  \textsc{TensorFlow Probability MCMC} \citep{lao2020tfp} in our simulations and the \textsc{ggplot2} package to generate all figures \citep{ggplot}.   In \textsc{R}, we use the package \textsc{coda} \citep{coda} to calculate effective sample sizes.  In \textsc{Python}, we use a built in function from \textsc{TensorFlow Probability MCMC}.  The primary color palette comes from \citet{wes}.

>>>>

 \begin{figure}[!t]
 	\vspace{3em}
	\centering
	\includegraphics[width=0.9\linewidth]{mcmcIterations.pdf}
	\caption{Total number of oracle evaluations required for each of 2,000 quantum parallel MCMC (QPMCMC) iterations for standard multivariate normal targets of five different dimensionalities. Regardless of target dimension, the individual QPMCMC runs require roughly $7$\% of the usual 4 million target evaluations. Over  99.4\% of the 10,000 MCMC iterations across all dimensions successfully sample from the discrete distribution with probabilities of \eqref{eq:probs}. Burn-in iterations require moderately more evaluations because the current state occupies a lower density region and represents a `less good' warm-start.}\label{fig:mcmcIterations}
\end{figure}

@@

In general MCMC, one often calibrates the scaling of a proposal kernel to balance between exploring the target's domain and remaining within high-density regions.  Optimal scaling strategies may lead to a large number of rejected proposals \citep{roberts2001optimal}.  Indeed, \citet{holbrook2021generating} shows that parallel MCMC algorithms are no exception. When using the Gumbel-max trick to sample from proposals, this means that the current state is often quite near optimality, representing a warm-start.  Figure \ref{fig:mcmcIterations} shows how this warm-starting effects the number of oracle calls (and hence target evaluations) within the quantum minimization sub-routine over the course of an MCMC trajectory.  We target multi-dimensional standard normal distributions of differing dimensions using the vector $(100,\dots,100)$ as starting position. We fix the number of Gaussian proposals to be 2,000 and use standard adaptation procedures \citep{rosenthal2011optimal} to target a 50\% acceptance rate while guaranteeing convergence. Although no theoretical results validate this target acceptance rate, the rate is close to empirically optimal rates from \citet{holbrook2021generating}. Across iterations, QPMCMC requires relatively few oracle calls compared to the 2,000 target evaluations of parallel MCMC. We also witness the expected drop in the number of oracle evaluations as the chain approaches the target's high-density region. Remarkably, the algorithm succeeds in sampling from the true discrete distribution in 99.5\% of the MCMC iterations. An independent simulation obtains similar results, requiring roughly 7\% of the usual number of target evaluations. Figure \ref{fig:qq} shows a quantile-quantile plot for all 100 dimensions of a multi-dimensional standard normal distribution. We see no appreciable impact from the rare failure of the quantum minimization sub-routine.

@@

\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}
\ra{1.2}
\begin{table}[t!]
	\centering
	\resizebox{\textwidth}{!}{
		\begin{tabular}{@{}lllll@{}}\toprule
			Proposals & MCMC iterations  & Target evaluations & Speedup  & Efficiency gain  \\   \midrule
			1,000 & 249,398 (200,998,  311,998) & 24,988,963 (20,149,132, 31,265,011) & 9.98  (9.98, 9.98) & 1  \\   
			5,000 & 14,398 (12,998, 16,998) & 3,314,560 (2,989,418,  3,916,281) & 21.72  (21.70, 21.74) & 7.58  (6.25, 9.71) \\   
			10,000 & 5,998  (4,998,  6,998) & 1,993,484  (1,662,592,  2,330,842) & 30  (29.96, 30.26) & 12.87  (8.64, 18.80) \\    \bottomrule
		\end{tabular}
	}
	\caption{Racing to a minimum (between the two dimensions) of 100 effective samples for a target with 1,000 disjoint modes. `Speedup' is ratio between target evaluations required for sequential and quantum implementations. `Efficiency gain' is ratio between target evaluations required for 1,000 proposal and 5,000/10,000 proposal implementations. We report means (minima, maxima) across 5 independent runs.}\label{tab:res}
\end{table}

@@

\section{Note on simulations}\label{sec:sim}

We use \textsc{R} \citep{rlang}, \textsc{Python} \citep{vanrossum1995python}, \textsc{TensorFlow} \citep{abadi2016tensorflow} and  \textsc{TensorFlow Probability MCMC} \citep{lao2020tfp} in our simulations and the \textsc{ggplot2} package to generate all figures \citep{ggplot}.   In \textsc{R}, we use the package \textsc{coda} \citep{coda} to calculate effective sample sizes.  In \textsc{Python}, we use a built-in function from \textsc{TensorFlow Probability MCMC}.  The primary color palette comes from \citet{wes}.

### 2023-01-25 manuscript/revision_letter.tex
just submitted revision

\begin{reply}
	Sometimes Algorithm \ref{alg:min} does not find the minimum.  Because of this the QPMCMC chain does not have the same transition kernel as the non-quantum version.  Nonetheless, results like those in Figure \ref{fig:qq} show that bias remains small.  The second paragraph of Section \ref{sec:disc} reads:
	\begin{quote}
This work leads to three interesting questions.   First, what is the status of inference obtained by QPMCMC?  The QPMCMC selection step relies on quantum minimization, an algorithm that only achieves success with probability $1-\epsilon$.  While empirical studies suggest that this error induces little bias, it would be helpful to use this $\epsilon$ to bound the distance between the target distribution and the distribution generated by QPMCMC.  Such theoretical efforts would need to extend recent formalizations of the multiproposal based parallel MCMC paradigm \citep{glatt} to account for biased MCMC kernels.
	\end{quote}
I am currently working with collaborators to bound the Wasserstein distance between the target and the distribution actually sampled.  At the moment, I must leave such results for future work. 
\end{reply}

@@

@@

@@

@@

@@

\begin{itemize}	
	\item Could the authors comment on the potential applicability of the algorithm on discrete state spaces? This may be useful for Bayesian record linkage, and variable selection problems \citep{zanella2020informed}.

>>>>

\begin{reply}
	Sometimes Algorithm \ref{alg:min} does not find the minimum. For example, in the experiments described in Figures \ref{fig:mcmcIterations} and \ref{fig:qq}, the correct kernel is used 99.4\% of the time.  Because of this the QPMCMC chain does not have the same transition kernel as the non-quantum version once every 200 QPMCMC iterations.  Nonetheless, results like those in Figure \ref{fig:qq} show that bias remains small.  The second paragraph of Section \ref{sec:disc} reads:
	\begin{quote}
This work leads to three interesting questions.   First, what is the status of inference obtained by QPMCMC?  The QPMCMC selection step relies on quantum minimization, an algorithm that only achieves success with probability $1-\epsilon$.  While empirical studies suggest that this error induces little bias, it would be helpful to use this $\epsilon$ to bound the distance between the target distribution and the distribution generated by QPMCMC.  Such theoretical efforts would need to extend recent formalizations of the multiproposal based parallel MCMC paradigm \citep{glatt} to account for biased MCMC kernels.
	\end{quote}
I am currently working with collaborators to bound the Wasserstein distance between the target and the distribution actually sampled.  At the moment, I must leave such results for future work. 
\end{reply}

@@

\begin{reply}
	I thank the AE for pointing out this error on my part.  In fact, the first line should have included `quantum parallel MCMC' instead of just `parallel MCMC'.

	\begin{quote}
		Total number of oracle evaluations required for each of 2,000 quantum parallel MCMC (QPMCMC) iterations for standard multivariate normal targets of five different dimensionalities. Regardless of target dimension, the individual QPMCMC runs require roughly $7$\% of the usual 4 million target evaluations. Over  99.4\% of the 10,000 MCMC iterations across all dimensions successfully sample from the discrete distribution with probabilities of \eqref{eq:probs}. Burn-in iterations require moderately more evaluations because the current state occupies a lower density region and represents a `less good' warm-start.
	\end{quote}
\end{reply}

@@

\begin{reply}
	I completely agree with the AE on this one.  The reference to \citet{rosenthal2011optimal} was for the `standard adaptation procedure', and not the 50\% tuning target.  I use this target because it is close to the roughly 60\% target I found optimal in my paper about another multiproposal technique \citep{holbrook2021generating} and because there are no theoretical optimality results in the multiproposal setting that I am aware of.  I've edited the text to make this clear:
	\begin{quote}
		We fix the number of Gaussian proposals to be 2,000 and use standard adaptation procedures \citep{rosenthal2011optimal} to target a 50\% acceptance rate while guaranteeing convergence. Although no theoretical results validate this target acceptance rate, the rate is close to empirically optimal rates from \citet{holbrook2021generating}.
	\end{quote}
\end{reply}

@@

\begin{reply}
For that and other related examples, I found that roughly 1 in 200 implementations of Algorithm 4 failed to find the minimum, i.e., failed to use the theoretically justified kernel.  We can make this failure happen less often by running Algorithm 4 longer, but there is a tradeoff between accuracy and target evaluations. Empirical results like those of Figure \ref{fig:qq} show that bias is relatively small, but I am currently working with collaborators to better understand how intermittent failure leads to biased sampling.

I quantify these things by running Monte Carlo simulations of all nested quantum algorithms beginning with the Grover iteration as basic building block.  This is implemented in R as described in Section \ref{sec:sim}.
\end{reply}

@@

\begin{reply}
	I include burnin here in order to demonstrate that warm-starting Algorithm 4 with the current Markov chain state becomes more beneficial as one converges the the target's high density regions. This explains why more target evaluations are required earlier on in the trajectory.
\end{reply}

@@

\begin{reply}
	I thank the AE for pointing out this omission. Here, I am using the minimum ESS between the two dimensions.  I  have edited the first sentence of the caption to say:
	\begin{quote}
		Racing to a minimum (between the two dimensions) of 100 effective samples for a target with 1,000 disjoint modes. 
	\end{quote}
 In case this was a software question, I have also added information on ESS to the first paragraph of Section \ref{sec:sim}:
 \begin{quote}
We use \textsc{R} \citep{rlang}, \textsc{Python} \citep{vanrossum1995python}, \textsc{TensorFlow} \citep{abadi2016tensorflow} and  \textsc{TensorFlow Probability MCMC} \citep{lao2020tfp} in our simulations and the \textsc{ggplot2} package to generate all figures \citep{ggplot}.   In \textsc{R}, we use the package \textsc{coda} \citep{coda} to calculate effective sample sizes.  In \textsc{Python}, we use a built-in function from \textsc{TensorFlow Probability MCMC}.  The primary color palette comes from \citet{wes}.
 \end{quote}
\end{reply}

\begin{itemize}	
	\item Could the authors comment on the potential applicability of the algorithm on discrete state spaces? This may be useful for Bayesian record linkage, and variable selection problems \citep{zanella2020informed}.
\end{itemize}

\begin{reply}
	This is a great question. I thank the AE for inspiring some additions to this manuscript.  I have added experiments in which I used QPMCMC to sample from an Ising model and a Bayesian classification model.  Figure \ref{fig:ising2d} shows benefits for using many proposals within QPMCMC and that speedups increase with the number of proposals!
\end{reply}

\setcounter{figure}{6}
 \begin{figure}[!t]
	\centering
	\includegraphics[width=0.7\linewidth]{figures/Ising2dFig.pdf}
	\caption{Convergence of log-posterior for a parallel MCMC sampler targeting an Ising model with uniform interaction $\rho=1$ and no external field. All chains begin at minimal probability `checkerboard' state. Larger proposal counts allow discovery of higher probability configurations.}\label{fig:ising2d}
\end{figure}

\begin{reply}
	Again, I thank the AE for all their helpful suggestions, and I am confident that the manuscript is now significantly improved as a result.
\end{reply}
